Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.00

### 2019-02-25 manuscript.Rmd
finished incorporating TC edits

## Custom gene and tissue expression boxplots
The 'Expression' tab of the webpage provides a wealth of information about both gene and transcript expression in a range of eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`). The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues. A boxplot is then generated afer hitting the "Re(Draw) Plot" button with overlaid individual data points. On mouse-over, the metadata for the individual sample is displayed. A tabular report is generated based on selected genes and tissues: a table with links to Ensembl, GeneCard, and OMIM for each gene for quick referencing, and a table containing expression levels for each selected gene in each selected tissue. The tables can be arranged or sorted to the user’s preference, and can be easily downloaded for local use.

```{r appScreenshots, echo = F, fig.cap = appScreenshots_cap, fig.width=12}
appScreenshots_cap <- fig_cap(name='appScreenshots', caption = 'Various screenshots from eyeIntegration web app. A. Pan-tissue gene expression box plots with accompanying data tables. B. Alternative heatmap visualization. C. CRX gene expression heatmap and table information from Macosko et al. P14 scRNA-seq. D. t-SNE visualization of individual cell gene expression profiles from Macosko et al. E. Data table export view. F. Differential gene expression across different tissues and with GO term enrichment. ')

@@

## Differential expression and gene ontology enrichment tests allow quick comparison of gene differences between groups
Using our expression data, we performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in our subset of the GTEx dataset, allowing users to quickly identify eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons. We have expanded the differential tests in the 2019 EiaD by adding the GTEx tissues as direct comparisons to our eye sub-tissues. The user can view the results under the ‘Differential’ selection under the ‘expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).  As with expression, the user can select which version of the web app to draw data from as well as select for gene or transcript level comparisons. The user additionally has the option to select different gene classes to examine, protein coding, lincRNA etc.

The results of differential expression are presented in a tabular format showing log2 fold change, average expression, and different statistical values. Depending on the comparison, there are `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% head(1)` to `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% tail(1)` differentially expressed genes (`r supTab_cap(name = 'limmaTests', display = 'cite')`). The table can be easily searched for any given gene, viewed and ordered to the user’s preference, and downloaded in CSV format. Differential expression can be visualized through fold change bar graphs in the ‘pan-tissue plots’ section under expression. Additionally, we performed GO enrichment for all differential comparisons. Enriched GO terms are presented first as a word cloud, for quick comparison of GO enrichment.  We provide tables, with similar viewing options as the differential expression table, for enriched GO terms in each class of a given differential comparison.

## Murine scRNA-seq enables testing of retina cell type specific expression
We provide scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a]. This allows researchers to quickly examine gene expression across individual cell types in the retina. Single cell gene expression data is visualized through a heatmap showing the expression of a gene across multiple retinal cell types and different developmental time points (E11-P14, if available), and a table of expression values is generated containing the expression data used to draw the heatmap (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'C')`). We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'D, see Methods')`). The plots show all cell types present a given developmental stage, and highlights cells expressing a gene above a user given level.

## EiaD 2019 demonstrates that organoids, stem cell-derived, and fetal tissue retina have closely related transcriptomes
There are, currently, two major approaches to studying developing human retina: post-mortem fetal tissue and stem-cell based organoids. We looked at whether at how well these approaches to studying developing retina compare at a transcriptomic level, both for tissue - organoid relationships and how well they correlate across early development.

@@

## Data accessibility 
Individual data files for gene expression and sample metadata can be downloaded from the ‘Data’ tab on the web app. All data and code used to generate the web app can be installed from the R command line by running devtools::install_github(‘davidmcg/eyeIntegration_app’). Code for the EiaD data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.

@@

EiaD 2019 contains a large set of carefully curated, reproducibly processed human eye RNA-seq datasets alongside a crucial human body tissue comparison set from the GTEx project. It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and is served via a powerful web app, eyeIntegration at https://eyeIntegration.nei.nih.gov. The web app offers a wide range of user-driven visualizations to compare expression of genes across dozens of human body and eye tissues. Furthermore, murine scRNA-seq datasets have been integrated, allowing for examination of retina cell type specific gene expression. Several human and non-human primate studies have been posted in the past year on the bioRxiv pre-print server. As the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].

As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for stem cell and organoid based developmental methods to be developed. Our merging of these datasets and analysis at a transcriptome level (as compared to evaluating with a few marker genes) demonstrates that these two approaches are successfully recapitulating fetal tissues to a first approximation at the whole transcriptome level. This result underscores the value of these approaches in building models for human retina developmental .

@@

pca_cap <- supFig_cap(name='allSamplePCA', caption = 'PCA plot of all samples demonstrates non-eye tissue most similar to adult retina is the brain, pituitary most similar to adult RPE, and skin and blood similar to cornea.')

>>>>

## Custom gene and tissue expression boxplots
The 'Expression' tab of the webpage provides a wealth of information about both gene- and transcript-level expression for eye and body tissues, giving the user the ability to compare the expression of different genes within a single tissue, as well as the expression of genes across multiple tissues  (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'A')`). The user first selects either the 2017 or 2019 gene or transcript EiaD dataset, then genes (or transcripts), then tissues. A boxplot is then generated after hitting the "Re(Draw) Plot" button with overlaid individual data points. On mouse-over, the metadata for the individual sample is displayed. A tabular report is generated based on selected genes and tissues: a table with links to Ensembl, GeneCard, and OMIM for each gene for quick referencing, and a table containing expression levels for each selected gene in each selected tissue. The tables can be arranged or sorted to the user’s preference, and can be easily downloaded for local use.

```{r appScreenshots, echo = F, fig.cap = appScreenshots_cap, fig.width=12}
appScreenshots_cap <- fig_cap(name='appScreenshots', caption = 'Screenshots from eyeIntegration web app. A. Pan-tissue gene expression box plots with accompanying data tables. B. Heatmap visualization. C. CRX gene expression heatmap and table information from Macosko et al. P14 scRNA-seq. D. t-SNE visualization of gene expression profiles of individual cells from Macosko et al. E. Data table export view. F. Differential gene expression across different tissues and with GO term enrichment.')

@@

## Differential expression and gene ontology enrichment tests allow quick comparison of gene differences between groups
We performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in our subset of the GTEx dataset, allowing quick identification of eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons. We have expanded the differential tests in the 2019 EiaD by adding the GTEx tissues as direct comparisons to our eye sub-tissues. The user can view the results selecting ‘Differential’ under the ‘Expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).  As with 'Expression', the user can select which version of the web app to draw data from as well as select for gene- or transcript-level comparisons. The user additionally has the option to select different gene classes to examine, e.g. protein coding, lincRNA.

The results of differential expression are presented in a tabular format showing log2 fold change, average expression, and different statistical values. Depending on the comparison, there are `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% head(1)` to `r q01_limma_gene_sig %>% pull(Count) %>% sort() %>% tail(1)` differentially expressed genes (`r supTab_cap(name = 'limmaTests', display = 'cite')`). The table can be easily searched for any given gene, viewed and ordered to the user’s preference, and downloaded in CSV format. Differential expression can be visualized through fold change bar graphs with the ‘Pan-tissue plots’ selection under ‘Expression’. Additionally, we performed GO enrichment for all differential comparisons. Enriched GO terms are presented first as a word cloud, for quick comparison of GO enrichment.  We provide tables, with similar viewing options as the differential expression table, for enriched GO terms in each class of a given differential comparison.

## Murine scRNA-seq enables testing of retina cell type specific expression
We incorporated scRNA-seq data from murine retina across two studies [@clarkComprehensiveAnalysisRetinal2018; @macoskoHighlyParallelGenomewide2015a]. This allows researchers to quickly examine gene expression across individual cell types in the retina. Single cell gene expression data is visualized through a heatmap showing the expression of a gene across multiple retinal cell types and different developmental time points, from embryonic day (E)11 to postnatal day (P)14 (when available), and a table of expression values is generated containing the expression data used to draw the heatmap (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'C')`). We also provide t-SNE/UMAP based clustering using cell type specific labeling created by the publishing authors (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'D, see Methods')`). The plots show all cell types present at a given developmental stage, and highlights cells expressing a gene above a user-selected given level.

## EiaD 2019 demonstrates that iPSC-derived organoids and fetal retina have closely related transcriptomes
There are, currently, two major approaches to studying developing human retina: post-mortem fetal tissue and stem-cell based organoids. We looked at whether at how well these approaches to studying developing retina compare at a transcriptomic level, both for tissue - organoid relationships and how well they correlate across early development.

@@

## Data accessibility 
Individual data files for gene expression and sample metadata can be downloaded from the ‘Data’ tab on the web app. All data and code used to generate the web app can be installed from the R command line by running devtools::install_github(‘davidmcg/eyeIntegration_app’). The code for the EiaD data processing pipeline can be found at https://github.com/davidmcg/Eyeintegration_autobuild.

@@

EiaD 2019 contains a large set of carefully curated, reproducibly processed human eye RNA-seq datasets alongside a human body tissue comparison set from the GTEx project. It is available for local install as an R package at https://www.github.com/davemcg/eyeIntegration_app and it is served via a web app, eyeIntegration at https://eyeIntegration.nei.nih.gov. The web app offers a wide range of user-driven visualizations to compare expression of genes across dozens of human body and eye tissues. Furthermore, murine scRNA-seq datasets have been incorporated, allowing for examination of retina cell type-specific gene expression. Several human and non-human primate studies have been posted in the past year on the pre-print server bioRxiv and as the raw data becomes publicly available, we will be updating this section of eyeIntegration[@lukowskiGenerationHumanNeural2018; @luSinglecellRNAseqAnalysis2018; @pengMolecularClassificationComparative2018].

As human fetal tissue is difficult to obtain and thus not very amenable for chemical or genetic modification, it is crucial for iPSC- and organoid-based in vitro developmental models to be developed. Our merging of these datasets and analysis at the transcriptome level (as compared to cross-analyzing using a limited number of known marker genes) demonstrates that these two approaches successfully recapitulate fetal tissues to a first approximation at the whole transcriptome level. This result underscores the value of these approaches in building models for human retina development.

@@

pca_cap <- supFig_cap(name='allSamplePCA', caption = 'PCA plot of all samples suggests that the non-eye tissue most similar to adult retina is the brain, pituitary is the most similar to adult RPE, and skin and blood are similar to cornea.')
